Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information

### 2022-03-22 write-up/letters/response-ref-1-v1.tex

	We clarify that this notation was chosen to match standard survey statistics notation: $Y_i$ is the true, fixed value for individual $i$ from the population, while $y_i$ is the observed value for individual $i$ from the sample. To help the reader, we now include a notation glossary in Appendix B.
	\vspace{5mm}

	\item {\it The “natural candidate for AIR” presented is the quantity that has more commonly been referred to as the “test positivity rate” (assuming that yi does in fact indicate test status rather than infection status which is obviously unknown). It would be helpful to mention this.}
	\vspace{5mm}

	This is now stated explicitly in Section 3, paragraph 2.
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 3.1: Imperfect testing}
\begin{enumerate}
	\item {\it Here, the value of distinguishing between test status and infection status becomes clearer. By using $P_j$ to mean ‘error status of testing’, the derivations in this section lead to a focus on false positives and false negatives. However, the real quantity of interest both for individuals and for determining prevalence is surely the positive predictive value and the negative predictive value – that is, Pr($Y_j$=1|positive test result) and Pr($Y_j$=0|negative test result). Unlike the FP and FN which (because they are strictly functions of sensitivity and specificity) are only affected by test characteristics, the PPV and NPV are affected also by prevalence. Acknowledging this leads more directly to the benefit of an iterative method for validating the test results (since FP \& FN should not depend on Y or y \& thus not need an iterative method).}
	\vspace{5mm}

	The adjustment is exactly related to the Bayes rule PPV/NPV relationship with FP, FN, and $\bar Y$.
	\textcolor{red}{I'm still a bit confused by this point.  Will review further.}
	\vspace{5mm}
	\item {\it Also worth noting that test positivity is not strictly binary (since it’s generally based on an underlying continuous cycle threshold), and so measurement error could potentially lead towards bias away from, or across, the null even in the absence of selection bias.}
	\vspace{5mm}

@@

	\item {\it The estimator $y = (\bar y_n-FP) /(1-(FP+FN))$ is a standard estimator for correcting for measurement error using FP \& FN estimated from a validation sample, where yn is the mis-measured average \& y is the corrected average. The use of this estimator to iteratively estimate the point prevalence is a good contribution of this paper, but the failure to distinguish between infection \& test status in the variable names somewhat obscures that this is what the author is doing.}
	\vspace{5mm}

	We now more carefully distinguish this and connect with the epidemiologic terminology to help readers from epidemiology.  And yes, this is the standard estimator.  We clarify, as you stated above, that the contribution of this paper was to derive this estimator from a new perspective (i.e., the decomposition perspective which didn't require any Bayesian machinery).
	\vspace{5mm}

	\item {\it Finally, in the section on effective sample size, please clarify why the FP and FN would be dependent on testing capacity. False positive and false negative rates, as well as sensitivity and specificity, are strictly characteristics of a given test – if increased test capacity or burden would alter the sample collection quality then FP \& FN could change. But if increased test capacity simply changes the underlying prevalence in the group tested (because of opening up testing to uninfected individuals), then the FP \& FN should not change, but the PPV and the NPV will change since these are dependent on prevalence.}
	\vspace{5mm}

	We clarify that increasing testing capacity may impact FP and FN rates if altering changes the sample collection quality, e.g., a testing center switches from nasal swab to oropharyngeal swabs or saliva specimen to increase testing capacity will alter measurement error due to differences in the test characteristics.
\end{enumerate}
\item {\bf Section 3.2: Regrettable Rates}
\begin{enumerate}
	\item {\it Here, and throughout the theoretical example, the model is an SIR model. However, we know that SARS-CoV-2 is better represented with an SEIR model, and indeed the applied example later in the manuscript uses an SEIR approach. Since this paper is very particularly addressing an issue with COVID data it would be beneficial to use an SEIR model throughout rather than the simpler SIR model.}
	\vspace{5mm}

	The SIR model in Section 3.2 is replaced by the SEIR model.  We highlight that the SEIR model has been shown to better represent the SARS-CoV-2 dynamics.  Figure 3a presents updated versions of the potential bias in ratio estimators under an SEIR model.   The conclusion remains largely the same. The only difference from the prior manuscript is that the peak in bias prior to the ratio going less than 1 is less pronounced (prior submission with SIR model has an approximate peak of 5 while current manuscript has approximate peak of 2).
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 3.3: Effective Reproduction Number}
\begin{enumerate}
	\item {\it Here again, an SEIR model would be valuable. It is also important to clarify that the effective reproductive number can be defined in a number of subtly different ways, and that these impact both the method of estimation and the interpretation of the effect. See for example Cori et al; Am J Epi, 178(9):1505-12. 2013 and Wallinga \& Tuenis Am J Epi; 160(6):509-16. 2004 for methods that, while similar, estimate subtly different quantities. It would be helpful to clarify exactly what effective reproductive number is being estimated here – from the formulae, I suspect it is the type described by Cori et al rather than Wallinga \& Tuenis.}
	\vspace{5mm}

@@

	We clarify the subtle differences between the definitions of the instantaneous and case reproductive numbers.  We establish that our approach aims to estimate the instantaneous reproduction number, i.e., is closely related to the proposal by Cori et al. (2013)).

	\vspace{5mm}
	\item {\it The Rt estimator will differ if an SEIR model is used, and it is not clear (to me) that the same intuition about the error will hold.}
	\vspace{5mm}

	%https://academic.oup.com/aje/article/189/11/1438/5864525?searchresult=1, plot SIR
	%https://academic.oup.com/aje/article/178/9/1505/89262?searchresult=1, CORI
	%https://www.nature.com/articles/s41598-020-76563-8.pdf?origin=ppub, Althaus approximate

	This estimator is originally motivated by the SIR model and is exact under the poisson likelihood; however, the moment-based estimator has been proposed as a general tool for assessing epidemic trends in real time (Xinhua Yu, Assessing Epidemic Trends in Real Time With a Simple Ratio Plot, Am J of Epid, 189(11), 2020).  The estimator is closely related to the instantaneous reproduction numbers proposed by Cori et. al. (2013), but we now rely on recent work (Heng and Althaus, The approximately universal shapes of epidemic curves in the SEIR model, Sci Reports, 2020) to propose an approximate moment-based estimator in the SEIR setting. Specifically, Heng and Althaus (2020) derive the following approximate formula for the basic reproduction number under SEIR dynamics:
	$$
	R_0 = \frac{1 + \Lambda \left(\frac{1}{\sigma} + \frac{1}{\gamma} \right)}{S_0 + I_0 (1 + \gamma/\sigma)}.
	$$
	where~$\Lambda$ is the epidemic growth rate, and $S_0$/$I_0$ are the initial fractions of the population that are susceptible and infectious respectively.  For $S_0 \approx 1$ and $I_0 \ll 1$, we have the approximate relationship $R_0 \approx 1 + (\frac{1}{\gamma} + \frac{1}{\sigma})\Lambda$.
	, this is an approximate solution which lacks the term $\Lambda^2 \frac{1}{\gamma \sigma}$.  It corresponds to the case of an exponentially distributed generation time with mean duration $1/\gamma + 1/\sigma$, which is the same as the solution for the SIR model assuming an infectious period of $1/\gamma + 1/\sigma$. For realistic growth rates and choices of $\sigma$ and $\gamma$ for the COVID-19 pandemic, this term is negligible and therefore we work with the approximate formula.
	\vspace{5mm}

	\noindent Note that for the SIR model, the basic reproduction number is~$R_0 = 1 + \frac{1}{\gamma} \Lambda$. So estimate of the instantaneous reproduction number is obtained by replacing $\Lambda$ by an estimate of the  instantaneous growth rate $\hat \Lambda_t = \log (K_t / K_{t-1})$.  Then using the approximate relationship between the SEIR model and an SIR model with infectious period $1/\gamma + 1/\sigma$, we arrive at the new moment-based estimator
	$$
    R_t \approx 1 + \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{K_t}{K_{t-1}} \right).
	$$
	The updated $R_t$ estimator shows that the same intuition about the error holds when we are considering an SEIR rather than an SIR model.  Figure 3b shows the reproduction number estimates and potential biases under SEIR dynamics which shows the same general conclusions hold.  For completeness, we also present the instantaneous reproduction number estimates using the proposal of Cori et. al. (2014) in Section I of the supplementary materials, which shows the same general conclusions.
\end{enumerate}
\item {\bf Section 3.4: Rate comparisons}
\begin{enumerate}
	\item {\it This section would also be more relevant with an SEIR model}
	\vspace{5mm}

@@

	We apologize for the confusion.  We now include distinct notation for each of the sampling mechanisms.  To strike a balance and keep notation at a minimum, we decided to use superscripts $R$ and $NR$ to distinguish between the probability and non-probability samples.  To facilitate comprehension, we have added a notation glossary in Section B of the supplementary materials.
	\vspace{5mm}
	\item {\it In the final formula (eq 4.3), I am not clear on why the weights are applied to both yk and FP (ie why sum(w*(yk-FP)) and not sum(w*yk) – FP)? This is very likely just me not understanding the derivation in Appendix C, but some explanation of the intuition would be helpful.}
	\vspace{5mm}

	Since weights are normalized, these are equivalent, i.e., $\sum w (x_i) FP / \sum w(x_i) = FP$.
	\vspace{5mm}
	\item {\it The time-varying section seems somewhat unnecessary since it is simply taking the expectation of the data over some time period – the daily data also does this, but where the time-period is 1 day. }
	\vspace{5mm}

	Section 4.1.3 allows researchers to choose to go beyond simpler moving window averages (i.e., daily/weekly averages).  The kernel function allows the user to choose weighted averages that may be more robust than the uniform kernel (which is equivalent to the a moving average over some fixed window of time).  The proposed approach helps smooth the estimated propensity over time; moreover, the flexibility of the approach for government agencies who may observed random samples at different frequencies is of practical importance.
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 4.2: model-based estimation}
\begin{enumerate}
	\item {\it The SEIR model described in the text and in the supplement differ in a way that I find confusing – the code and equations describe a simple SEIR model but the written text describes an SEmInR model with stratum-specific covariates which appear to apply to at least the I compartments and possibly the E compartments. From the code and equations, presumably the assumption is that all strata mix with identical probability and all strata experience the same probabilities of infection, exposure, and transmission. However, these assumptions are (1) unlikely to hold and (2) make it hard to understand why the model includes strata at all. }
	\vspace{5mm}

### 2022-03-22 write-up/covid-supplement-writeup.tex
updated paper

Table~\ref{tab:comparison} demonstrates minimal bias in the Facebook symptom survey with respect to symptom distributions.

 \begin{table}[!th]
 \centering
 \begin{tabular}{c | c c c | c}
 & \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-4}
 & Lower CI & Estimate & Upper CI & Estimate \\ \hline
 Fever &  0.073 & 0.097 & 0.012 & 0.018 \\
 Cough &  0.140 & 0.150 & 0.159 & 0.149 \\
 Shortness & 0.055 & 0.062 & 0.068 & 0.062 \\ \hline
 \end{tabular}
 \caption{Comparison of Facebook survey and random sample}
 \label{tab:comparison}
 \end{table}

@@

Here we present the estimate of~\cite{Cori20113}.  Let~$I_t$ detnoe the total infectiousness of infected individuals at time~$t$.  Then~$E[I_t] = R_t \sum_{s=1}^{t} I_{t-s} w_s$ where $w_{u}$ is the infectivity function.  Here, we follow~\cite{Cori20113} and choose a discretized shifted Gamma distribution with mean $7$ days and standard deviation $2$ days.  Then a moment-based estimator can be obtained by $\hat R_t = I_t / \sum_{s=1}^t I_{t-s} w_s$, which takes the form of the ratio estimators in Section~\ref{section:rates}.  Therefore, the bias can be readily obtained from the associated Taylor series decomposition, with the terms related to time~$t-1$ replaced by a weighted version.

>>>>

Table~\ref{tab:comparison} demonstrates minimal bias in the Delphi’s COVID-19 Trends and Impact Survey (CTIS) with respect to symptom distributions.

 \begin{table}[!th]
 \centering
 \begin{tabular}{c | c c c | c}
 & \multicolumn{3}{c}{CTIS} & Random \\ \cline{2-4}
 & Lower CI & Estimate & Upper CI & Estimate \\ \hline
 Fever &  0.073 & 0.097 & 0.012 & 0.018 \\
 Cough &  0.140 & 0.150 & 0.159 & 0.149 \\
 Shortness & 0.055 & 0.062 & 0.068 & 0.062 \\ \hline
 \end{tabular}
 \caption{Comparison of CTIS and random sample restricting CTIS to surveys collected from April 25--29th, 2020.}
 \label{tab:comparison}
 \end{table}

@@

Here we present estimates of the instantaneous reproductive number using the approach of~\cite{Cori20113} under the SEIR model from Section~\ref{section:r0-estimation}.  Let~$I_t$ denote the total infectiousness of infected individuals at time~$t$.  Then~$E[I_t] = R_t \sum_{s=1}^{t} I_{t-s} w_s$ where $w_{u}$ is the infectivity function.  Here, we follow~\cite{Cori20113} and choose a discretized shifted Gamma distribution with mean $7$ days and standard deviation $2$ days.  Then a moment-based estimator can be obtained by $\hat R_t = I_t / \sum_{s=1}^t I_{t-s} w_s$, which takes the form of the ratio estimators in Section~\ref{section:rates}.  Therefore, the bias can be readily obtained from the associated Taylor series decomposition, with the terms related to time~$t-1$ replaced by a weighted version.  Figure~\ref{fig:cori-bias} presents the bias.

### 2022-03-22 write-up/covid-writeup.tex
updated paper

 Tests are imperfect.  COVID-19 testing is no exception. Here we investigate the interplay between imperfect testing and selection bias.  Researchers often assume measurement error leads to parameter attenuation.  When paired with selection bias, however, the two sources become entangled, and resulting errors in magnified, muted, or even switch signs.

@@

 \begin{remark}
 \label{rmk:testbinary}
 Note that test positivity is based on an underlying continuous cycle threshold and therefore not strictly binary.  While we focus on the interplay of measurement error and selection bias in the context of binary outcomes due to the dichotomous nature of the COVID-19 testing data, it should be noted that measurement error can lead to bias even in the absence of selection bias.
 \end{remark}

@@

 \begin{figure}[!th]
 \centering
 \includegraphics[width = 0.7\textwidth]{../figs/mem_heatmap.png}
 \caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.024$ and $FN=0.13$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
 %\textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
 \label{fig:heatmap}
 \vspace{-0.3cm}
 \end{figure}

 Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing, we see that
 $$
 \frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right].
 $$
 % implying the error relative to SRS increases as a function of population size~\citep{Meng2018}. That two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
 % % Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
 A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
 $$
 \frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
 $$
 Between April 25th to 29th 2020, Indiana performed $95,879$ tests.  Indiana's population is roughly $6.732$ million, so $f = 0.003$.  The active infection rate was estimated to be~$1.81\%$~\citep{Yiannoutsos2021} in this time interval. Recent studies have suggested 87\% sensitivity~\citep{Arevalo2020} and 97.6\% specificity~\citep{Cohen2020} are reasonable measurement error rates for RT-PCR tests. Supposing COVID-19 positive individuals are $1.5$ times more likely to get tested, then the effective sample size is $168$. Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.  Even if the relative sampling rate drops to $1.2$ and $f$ increases to $0.01$ then the effective sample size will increase to $1025$.  Thus the effective sample size even in optimistic scenarios is equivalent to a moderate random sample from the population.  Moreover, increased testing capacity may alter false positive and negative rates due to changes in sample collection quality, e.g., a testing center switches from nasal swab to oropharyngeal swabs or saliva specimen to speed up data collection.  Consider the case where $f$ increasing from $0.003$ to $0.01$ and is associated with the false negative rate rising from $13\%$ to $20$\%.  Then the effective sample size is $863$, representing a $5.1$ factor increase rather than the expected $6.1$ factor increase. See Section~\ref{section:effss} in the Supplementary Materials for additional effective sample size calculations which show that while~\citep{Meng2018} argues the relative error increases as a function of population size, our calculations indicate this is not true for fixed relative frequency~$f_1/f_0$.
 % \begin{wrapfigure}{r}{0.5\textwidth}

@@

 Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under a susceptible-exposed-infected-recovered (SEIR) model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
 \begin{align}
 \frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \label{eq:seir} \\
 \frac{\partial i_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
 \frac{\partial r_t}{\partial t} = - \gamma i_t. \nonumber
\end{align}
 % \begin{equation}
 % \label{eq:sir}
 % \frac{\partial s_t}{\partial t} = - \beta s_t i_t; \quad
 % \frac{\partial i_t}{\partial t} = \beta s_t i_t - \gamma i_t; \quad
 % \frac{\partial r_t}{\partial t} = - \gamma i_t
 % \end{equation}
 where $s_t, e_t, i_t$ and $r_t$ are the fraction of susceptible, exposed infected, and removed (recovered or deceased) individuals in the population at time $t$ respectively.  The SEIR model has been used extensively as a model for SARS-CoV-2 dynamics~\cite{Song2020}.  In terms of bias, the rate is overestimated prior to the peak in the fraction infected and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.  Such biases may impact policy making.  Overestimation pre-peak may give policy makers more leverage in proposing aggressive actions to reduce prevalence.  Underestimation post-peak puts pressure on policy makers to prematurely relax social distancing measures.  Estimates at the peak time appear to have minimal bias.
 %Note, however, that the standard errors will be large due to small effective sample sizes.

@@

 Here we assume access to a probability sample measuring the same set of covariates.  Let $I_j^{R}$ be an indicator that the individual was included in the probability sample and $W_j^R$ be the probability of inclusion for the $j$th individual. \cite{Chen2019} use a probability sample to construct a design-unbiased estimator of the second term
 \begin{equation}
 \label{eq:auxinfoprob}
 \sum_{j=1}^N I_j^R \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N I_j^R W_j^R \log ( 1 - \pi (X_j; \theta)).
 \end{equation}
 Expectation of~\ref{eq:auxinfoprob} with respect to the sampling design yields~\eqref{eq:propensity}. Solving~\eqref{eq:auxinfoprob} is done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} in the supplementary materials for details.  For ease of comprehension, a notation glossary is provided in Section~\ref{app:notation} in the the supplementary materials.

 \begin{remark}
 Weights built from selection propensities~$\{\pi(x_i; \theta)\}_{i=1}^n$ are common practice in the epidemiology literature.  Survey sampling and transportability weights require knowledge of the selection mechanism~\citep{10.1093/aje/kwx164,ColeStuart2010}, which is not available in our setting.  Instead, we require auxiliary information in the form of a random sample to estimate the probability of selection. A key difference from prior approaches is the need to estimate time-varying probabilities of selection, e.g., Figure 6 shows how the selection propensity of young individuals changes dramatically over the observation window.
 \end{remark}

@@

 \begin{lemma}[Variance of IPW estimator] \normalfont
 \label{lemma:ipw}
 % \textcolor{red}{Must check.}
The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ at a time~$t \in \{t_l^\prime\}$ are solutions to the following set of estimating equations
 $$
 \Phi_n (\eta_t) =
 \left (
 \begin{array}{c}
 \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
 \frac{1}{N \times \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right)} \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I^{NR}_{j,t_l^\prime} X_{j,t_l^\prime} - \sum_{j = 1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime}  \right] \\
 \frac{1}{n_{FP}} \sum_{i=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
 \frac{1}{n_{FN}} \sum_{i=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
 \end{array}
 \right ) = {\bf 0}
 $$
 where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{h} (|t - t_l^\prime|)$ and $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ where $\phi_n = \frac{\partial \Phi (\eta)}{\partial \eta}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
 \end{lemma}

>>>>

 Tests are imperfect.  COVID-19 testing is no exception. Here we investigate the interplay between imperfect testing and selection bias.  Researchers often assume measurement error leads to parameter attenuation.  When paired with selection bias, however, the two sources become entangled, and resulting errors can be magnified, muted, or even switch signs.

@@

 \begin{remark}
 \label{rmk:testbinary}
 Note that test positivity is based on an underlying continuous cycle threshold and therefore not strictly binary.  We focus on the interplay of measurement error and selection bias in the context of binary outcomes due to the dichotomous nature of the COVID-19 testing data.  Note that measurement error can lead to bias even in the absence of selection bias.
 \end{remark}

@@

 \begin{figure}[!th]
 \centering
 \includegraphics[width = 0.7\textwidth]{../figs/mem_heatmap.png}
 \caption{Imperfect testing adjustment~($D_M$) contour plot as a function of relative frequency $f_1/f_0$ (x-axis) and odds ratio (y-axis) for $FP=0.024$ and $FN=0.13$.}
 %\textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
 \label{fig:heatmap}
 \vspace{-0.3cm}
 \end{figure}

 Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing, we see that
 $$
 \frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right].
 $$
 % implying the error relative to SRS increases as a function of population size~\citep{Meng2018}. That two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
 % % Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
 A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
 $$
 \frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
 $$
 Between April 25th to 29th 2020, Indiana performed $95,879$ tests.  Indiana's population is roughly $6.732$ million, so $f = 0.003$.  The active infection rate was estimated to be~$1.81\%$~\citep{Yiannoutsos2021} in this time interval. Recent studies have suggested 87\% sensitivity~\citep{Arevalo2020} and 97.6\% specificity~\citep{Cohen2020} are reasonable measurement error rates for RT-PCR tests. Supposing COVID-19 positive individuals are $1.5$ times more likely to get tested, then the effective sample size is $168$. Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.  Even if the relative sampling rate drops to $1.2$ and $f$ increases to $0.01$ then the effective sample size will increase to $1025$.  Thus the effective sample size even in optimistic scenarios is equivalent to a moderate random sample from the population.  Moreover, increased testing capacity may alter false positive and negative rates due to changes in sample collection quality, e.g., a testing center switches from nasal swab to oropharyngeal swabs or saliva specimen to speed up data collection.  Consider the case where $f$ increasing from $0.003$ to $0.01$ and is associated with the false negative rate rising from $13\%$ to $20$\%.  Then the effective sample size is $863$, representing a $5.1$ factor increase rather than the expected $6.1$ factor increase. See Section~\ref{section:effss} in the Supplementary Materials for additional effective sample size calculations.  Note that while~\citep{Meng2018} argues the relative error increases as a function of population size, our calculations indicate this is not true when relative frequency~$f_1/f_0$ is held fixed.
 % \begin{wrapfigure}{r}{0.5\textwidth}

@@

 Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under a susceptible-exposed-infected-recovered (SEIR) model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
 \begin{align}
 \frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \label{eq:seir} \\
 \frac{\partial i_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
 \frac{\partial r_t}{\partial t} = - \gamma i_t. \nonumber
\end{align}
 % \begin{equation}
 % \label{eq:sir}
 % \frac{\partial s_t}{\partial t} = - \beta s_t i_t; \quad
 % \frac{\partial i_t}{\partial t} = \beta s_t i_t - \gamma i_t; \quad
 % \frac{\partial r_t}{\partial t} = - \gamma i_t
 % \end{equation}
 where $s_t, e_t, i_t$ and $r_t$ are the fraction of susceptible, exposed, infected, and removed (recovered or deceased) individuals in the population at time $t$ respectively.  The SEIR model has been used extensively as a model for SARS-CoV-2 dynamics~\cite{Song2020}.  In terms of bias, the rate is overestimated prior to the peak in the fraction infected and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.  Such biases may impact policy making.  Overestimation pre-peak may give policy makers more leverage in proposing aggressive actions to reduce prevalence.  Underestimation post-peak puts pressure on policy makers to prematurely relax social distancing measures.  Estimates at the peak time appear to have minimal bias.
 %Note, however, that the standard errors will be large due to small effective sample sizes.

@@

 Here we assume access to a probability sample measuring the same set of covariates.  Let $I_j^{R}$ be an indicator that the individual was included in the probability sample and $W_j^R$ be the probability of inclusion for the $j$th individual. \cite{Chen2019} use a probability sample to construct a design-unbiased estimator of the second term
 \begin{equation}
 \label{eq:auxinfoprob}
 \sum_{j=1}^N I_j^R \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N I_j^R W_j^R \log ( 1 - \pi (X_j; \theta)).
 \end{equation}
 Expectation of~\ref{eq:auxinfoprob} with respect to the sampling design yields~\eqref{eq:propensity}. Solving~\eqref{eq:auxinfoprob} is done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} and Section~\ref{app:notation} in the supplementary materials for details and a notation glossary respectively.

 \begin{remark}
 Weights built from selection propensities~$\{\pi(x_i; \theta)\}_{i=1}^n$ are common practice in the epidemiology literature.  Survey sampling and transportability weights require knowledge of the selection mechanism~\citep{10.1093/aje/kwx164,ColeStuart2010}, which is not available in our setting.  Instead, here we require auxiliary information in the form of a random sample to estimate the probability of selection.
 \end{remark}

@@

 \begin{lemma}[Variance of IPW estimator] \normalfont
 \label{lemma:ipw}
 % \textcolor{red}{Must check.}
The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat{FP}$, and $\hat{FN}$ at a time~$t \in \{t_l^\prime\}$ are solutions to the following set of estimating equations
 $$
 \Phi_n (\eta_t) =
 \left (
 \begin{array}{c}
 \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
 \frac{1}{N \times \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right)} \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I^{NR}_{j,t_l^\prime} X_{j,t_l^\prime} - \sum_{j = 1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime}  \right] \\
 \frac{1}{n_{FP}} \sum_{i=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
 \frac{1}{n_{FN}} \sum_{i=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
 \end{array}
 \right ) = {\bf 0}
 $$
 where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{h} (|t - t_l^\prime|)$ and $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ where $\phi_n = \frac{\partial \Phi (\eta)}{\partial \eta}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
 \end{lemma}
